---
title: "ProjectCheckIn2"
format: pdf
editor: visual
---

#in top right Git tab: pull, edit, commit, push.

##NFL data

```{r}
library(tidyverse)
library(nflreadr)

# Pick one season first
season_year <- 2013:2023 #seasons end at 2023 because merging contract data with other sports from recent seasons gets tricky with, depending on the sport; updated salaries, useable API's abd packages for getting data, and all sports have comfortable and reliable data through 2023. This range also gives us a full decade of data.

nfl_stats <- load_player_stats(               #player stats
  seasons = season_year,
  summary_level = "reg"
)
nfl_contracts <- load_contracts() #contract data

```

#filtering and cleaning NFL stats data and creating the metric for player evals

```{r}

#turn pasing epa, rushing epa, and receiving epa, into just EPA
nfl_perf <- nfl_stats %>%
  filter(position %in% c("QB", "RB", "WR", "TE")) %>%
  mutate(
    raw_performance = coalesce(passing_epa, 0) +         #coalesce creates a 0 instead of NA for non passers
      coalesce(rushing_epa, 0) +
      coalesce(receiving_epa, 0)
  ) %>%
  select(
    player_id,
    player = player_display_name,
    season,
    team = recent_team,
    position,
    games,
    raw_performance
  )

```

#now clena and filter nfl contract data

```{r}
nfl_salary <- nfl_contracts %>%
  filter(!is.na(gsis_id)) %>%       #take only players with valid IDs or it will be hard to merge
  group_by(gsis_id) %>%             #group contracts of the same player to avoid repeates
  slice_max(apy, n = 1, with_ties = FALSE) %>% #keeps only the largest salay for each player, renegotiations and rookie deals otherwies effect the data and the highest value often is more representative of how a team values a player
  ungroup() %>%
  transmute(
    player_id = gsis_id,             #transmute to change var names
    salary = apy,
    team_contract = team
  )

```

#join nfl contracts and player stats

```{r}
nfl_all_with_salary <- nfl_stats %>%
  left_join(nfl_salary, by = "player_id")
```

#Final cleaning and adding salary share and team payroll variables

```{r}
nfl_all_with_payroll <- nfl_all_with_salary %>%
  filter(!is.na(salary)) %>%                         #remove missing slaries
  group_by(season, recent_team) %>%
  mutate(
    team_payroll = sum(salary, na.rm = TRUE)         #create team payroll before removing non-skill position players so when we create the salary_share variable later, it is representative of their share of the whole teams salary, nt jsut skill position
  ) %>%
  ungroup()

nfl_cleaned <- nfl_all_with_payroll %>%
  filter(position %in% c("QB", "RB", "WR", "TE")) %>% #filtering to skill positions because EPA is more directly attributable to them as opposed ot kickers of offensive linemen
  mutate(
    raw_performance =
      coalesce(passing_epa, 0) +
      coalesce(rushing_epa, 0) +
      coalesce(receiving_epa, 0),
    salary_share = salary / team_payroll,
    league = "NFL",
    performance_metric = "Total Offensive EPA"
  ) %>%
  select(
    league,
    player_id,
    player = player_display_name,
    season,
    team = recent_team,
    position,
    games,
    raw_performance,
    salary,
    team_payroll,
    salary_share,
    performance_metric
  )
```

##Baseall data

```{r}
library(baseballr)

seasons <- 2013:2023 #create our range of season data we are using

mlb_batting_raw <- map_dfr(seasons, function(year) {   #repeated the function for each season
  fg_bat_leaders(                                      #the actual function that pulls batter data for the selected season
    startseason = year,
    endseason = year
  )
})

mlb_pitching_raw <- map_dfr(seasons, function(year) {  #same for pitchers
  fg_pitch_leaders(
    startseason = year,
    endseason = year
  )
})
```

#Clean data and change variable names to match NFL data

```{r}
mlb_batting_clean <- mlb_batting_raw %>%        #mlb data will have duplicates because players get raded midseason 
  transmute(
    player_id = playerid,
    player = PlayerName,
    season = Season,
    team = team_name,
    position_group = "Batter",
    raw_performance = WAR,
    performance_metric = "WAR"
  )

mlb_pitching_clean <- mlb_pitching_raw %>%
  transmute(
    player_id = playerid,
    player = PlayerName,
    season = Season,
    team = team_name,
    position_group = "Pitcher",
    raw_performance = WAR,
    performance_metric = "WAR"
  )

mlb_clean <- bind_rows(
  mlb_batting_clean,
  mlb_pitching_clean
) %>%
  group_by(player_id, player, season) %>%        #not including team in the grouping because otherwise players who got traded mid seasons would have multiple observations for each team they played for within the same year. This way our observational unit stay as one player season, combining their war totals from all teams in a given year which keeps it similar to the nfl data.
  summarize(
    raw_performance = sum(raw_performance, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    league = "MLB",
    performance_metric = "WAR"
  )

mlb_clean %>%
  count(player_id, season) %>%
  filter(n > 1)
```

#MLB Salary Data (Kaggle)

```{r}
library(tidyverse)
library(rvest)
library(janitor)

mlb_salary_data <- read_csv("mlb_salary_data.csv")
```

#Cleaning salary data

```{r}
mlb_salary <- mlb_salary_data %>%
  filter(Year >= 2013, Year <= 2023) %>% #Filter to selected years
  transmute( #adjust variable names
    season = Year,
    team = Team,
    player = Name,
    salary = parse_number(Salary)     #removing commas and dollar sign, converting to simple integer
  )
```

#Creating salary share for MLB

```{r}
mlb_salary <- mlb_salary %>%
  group_by(season, team) %>%
  mutate(
    team_payroll = sum(salary, na.rm = TRUE),
    salary_share = salary / team_payroll
  ) %>%
  ungroup()
```

#Combining MLB stats and salaries

```{r}
mlb_clean %>%
  select(player) %>%
  slice(1:20)

mlb_salary %>%
  select(player) %>%
  slice(1:20)
```

#Ensuring all player names are in the same format my removing accent marks

```{r}
library(stringi)

mlb_clean <- mlb_clean %>%
  mutate(
    player_clean = stri_trans_general(player, "Latin-ASCII")
  )

mlb_salary <- mlb_salary %>%
  mutate(
    player_clean = stri_trans_general(player, "Latin-ASCII")
  )
```

#Joining

```{r}
mlb_joined <- mlb_clean %>%
  left_join(
    mlb_salary,
    by = c("player_clean" = "player_clean", "season" = "season")
  )

#check quality
mlb_joined %>%             #initially missing salary data for players whom we have stats for for roughly 5,000 observations. This is because our stats data includes every single player who appeared where the salary data is slightly less reliable, likely not giving us certain fringe players or players who only played a couple games or most likely, minor league contracts. 
  summarize(
    total_players = n(),
    matched_salary = sum(!is.na(salary)),
    missing_salary = sum(is.na(salary))
  )

mlb_joined <- mlb_joined %>%     #Removing all players with unmatched salary and statistical data leaves us with just over 10,000 observations and still a clear picture of salary efficiency. 
  filter(!is.na(salary))
```

#Make final MLB data set and remove duplicate player column

```{r}
mlb_cleaned <- mlb_joined %>%
  select(
    player_id,
    player = player.x,
    season,
    team,
    raw_performance,
    league,
    performance_metric,
    salary,
    team_payroll,
    salary_share
  )

#duplicates check
mlb_cleaned %>%
  count(player_id, season) %>%
  filter(n > 1)

#still duplicates
mlb_cleaned %>%
  filter(player_id == 494, season == 2019)
#since WAR is the same but the salary is differing we can filter by the first observation for the higher salary
mlb_cleaned <- mlb_cleaned %>%
  group_by(player_id, player, season) %>%
  summarize(
    team = first(team),
    raw_performance = first(raw_performance),
    league = first(league),
    performance_metric = first(performance_metric),
    salary = sum(salary, na.rm = TRUE),
    team_payroll = first(team_payroll),
    salary_share = salary / team_payroll,
    .groups = "drop"
  )
```

#NBA data (basketball reference)

```{r}
#Scraping basketball reference for each season in our range of years
library(tidyverse)
library(rvest)
library(janitor)

nba_2013_raw <- read_html("https://www.basketball-reference.com/leagues/NBA_2013_advanced.html") %>%
  html_table(fill = TRUE)

nba_2014_raw <- read_html("https://www.basketball-reference.com/leagues/NBA_2014_advanced.html") %>%
  html_table(fill = TRUE)

nba_2015_raw <- read_html("https://www.basketball-reference.com/leagues/NBA_2015_advanced.html") %>%
  html_table(fill = TRUE)

nba_2016_raw <- read_html("https://www.basketball-reference.com/leagues/NBA_2016_advanced.html") %>%
  html_table(fill = TRUE)

nba_2017_raw <- read_html("https://www.basketball-reference.com/leagues/NBA_2017_advanced.html") %>%
  html_table(fill = TRUE)

nba_2018_raw <- read_html("https://www.basketball-reference.com/leagues/NBA_2018_advanced.html") %>%
  html_table(fill = TRUE)

nba_2019_raw <- read_html("https://www.basketball-reference.com/leagues/NBA_2019_advanced.html") %>%
  html_table(fill = TRUE)

nba_2020_raw <- read_html("https://www.basketball-reference.com/leagues/NBA_2020_advanced.html") %>%
  html_table(fill = TRUE)

nba_2021_raw <- read_html("https://www.basketball-reference.com/leagues/NBA_2021_advanced.html") %>%
  html_table(fill = TRUE)

nba_2022_raw <- read_html("https://www.basketball-reference.com/leagues/NBA_2022_advanced.html") %>%
  html_table(fill = TRUE)

nba_2023_raw <- read_html("https://www.basketball-reference.com/leagues/NBA_2023_advanced.html") %>%
  html_table(fill = TRUE)

#Cleaning each year
nba_2013_clean <- nba_2013_raw[[1]] %>%
  clean_names() %>%
  filter(player != "Player") %>%
  transmute(player, season = 2013, team, position = pos, games = as.numeric(g),
            raw_performance = as.numeric(ws), league = "NBA", performance_metric = "Win Shares")

nba_2014_clean <- nba_2014_raw[[1]] %>%
  clean_names() %>%
  filter(player != "Player") %>%
  transmute(player, season = 2014, team, position = pos, games = as.numeric(g),
            raw_performance = as.numeric(ws), league = "NBA", performance_metric = "Win Shares")

nba_2015_clean <- nba_2015_raw[[1]] %>%
  clean_names() %>%
  filter(player != "Player") %>%
  transmute(player, season = 2015, team, position = pos, games = as.numeric(g),
            raw_performance = as.numeric(ws), league = "NBA", performance_metric = "Win Shares")

nba_2016_clean <- nba_2016_raw[[1]] %>%
  clean_names() %>%
  filter(player != "Player") %>%
  transmute(player, season = 2016, team, position = pos, games = as.numeric(g),
            raw_performance = as.numeric(ws), league = "NBA", performance_metric = "Win Shares")

nba_2017_clean <- nba_2017_raw[[1]] %>%
  clean_names() %>%
  filter(player != "Player") %>%
  transmute(player, season = 2017, team, position = pos, games = as.numeric(g),
            raw_performance = as.numeric(ws), league = "NBA", performance_metric = "Win Shares")

nba_2018_clean <- nba_2018_raw[[1]] %>%
  clean_names() %>%
  filter(player != "Player") %>%
  transmute(player, season = 2018, team, position = pos, games = as.numeric(g),
            raw_performance = as.numeric(ws), league = "NBA", performance_metric = "Win Shares")

nba_2019_clean <- nba_2019_raw[[1]] %>%
  clean_names() %>%
  filter(player != "Player") %>%
  transmute(player, season = 2019, team, position = pos, games = as.numeric(g),
            raw_performance = as.numeric(ws), league = "NBA", performance_metric = "Win Shares")

nba_2020_clean <- nba_2020_raw[[1]] %>%
  clean_names() %>%
  filter(player != "Player") %>%
  transmute(player, season = 2020, team, position = pos, games = as.numeric(g),
            raw_performance = as.numeric(ws), league = "NBA", performance_metric = "Win Shares")

nba_2021_clean <- nba_2021_raw[[1]] %>%
  clean_names() %>%
  filter(player != "Player") %>%
  transmute(player, season = 2021, team, position = pos, games = as.numeric(g),
            raw_performance = as.numeric(ws), league = "NBA", performance_metric = "Win Shares")

nba_2022_clean <- nba_2022_raw[[1]] %>%
  clean_names() %>%
  filter(player != "Player") %>%
  transmute(player, season = 2022, team, position = pos, games = as.numeric(g),
            raw_performance = as.numeric(ws), league = "NBA", performance_metric = "Win Shares")

nba_2023_clean <- nba_2023_raw[[1]] %>%
  clean_names() %>%
  filter(player != "Player") %>%
  transmute(player, season = 2023, team, position = pos, games = as.numeric(g),
            raw_performance = as.numeric(ws), league = "NBA", performance_metric = "Win Shares")
```

#Combine all the years together

```{r}
nba_clean <- bind_rows(
  nba_2013_clean,
  nba_2014_clean,
  nba_2015_clean,
  nba_2016_clean,
  nba_2017_clean,
  nba_2018_clean,
  nba_2019_clean,
  nba_2020_clean,
  nba_2021_clean,
  nba_2022_clean,
  nba_2023_clean
)

nba_clean <- nba_clean %>%
  group_by(player, season) %>%
  filter(
    team %in% c("2TM", "3TM", "4TM") | n() == 1            # for players who played for multiple teams, remove team specific observations and keep the data for their season totals across teams.
  ) %>%
  ungroup()

#check for dupes
nba_clean %>%
  count(player, season) %>%
  filter(n > 1)
```

#nba salary data (kaggle data set)

```{r}
library(readr)
NBA_salary_data <- read_csv("NBA_salary_data.csv")
View(NBA_salary_data)

#filtering
nba_salary <- NBA_salary_data %>%
  filter(Year >= 2013, Year <= 2023) %>%
  transmute(
    player = Player,
    season = Year,
    team = Team,
    salary = Salary
  )

#create salary share variable
nba_salary <- nba_salary %>%
  group_by(season, team) %>%
  mutate(
    team_payroll = sum(salary, na.rm = TRUE),
    salary_share = salary / team_payroll
  ) %>%
  ungroup()

#same cleaning and joining as MLB data 
library(stringi)

nba_clean <- nba_clean %>%
  mutate(player_clean = stri_trans_general(player, "Latin-ASCII"))

nba_salary <- nba_salary %>%
  mutate(player_clean = stri_trans_general(player, "Latin-ASCII"))

nba_joined <- nba_clean %>%
  left_join(
    nba_salary,
    by = c("player_clean", "season")
  )

#checking match rate
nba_joined %>%
  summarize(
    total_players = n(),
    matched_salary = sum(!is.na(salary)),
    missing_salary = sum(is.na(salary))
  )

#remove duplicate columns for player names and team names that came from both daat sets
nba_cleaned <- nba_joined %>%
  filter(!is.na(salary)) %>%
  select(
    player = player.x,
    season,
    team = team.x,
    position,
    games,
    raw_performance,
    league,
    performance_metric,
    salary,
    team_payroll,
    salary_share
  )
#check duplicates
nba_cleaned %>%
  count(player, season) %>%
  filter(n > 1)
```

#NHL data from hockey reference

```{r}
# NHL performance data pulled teh same way as nba

library(tidyverse)
library(rvest)
library(janitor)

# Pull each year separately
nhl_2013_raw <- read_html("https://www.hockey-reference.com/leagues/NHL_2013_skaters.html") %>%
  html_table(fill = TRUE)

nhl_2014_raw <- read_html("https://www.hockey-reference.com/leagues/NHL_2014_skaters.html") %>%
  html_table(fill = TRUE)

nhl_2015_raw <- read_html("https://www.hockey-reference.com/leagues/NHL_2015_skaters.html") %>%
  html_table(fill = TRUE)

nhl_2016_raw <- read_html("https://www.hockey-reference.com/leagues/NHL_2016_skaters.html") %>%
  html_table(fill = TRUE)

nhl_2017_raw <- read_html("https://www.hockey-reference.com/leagues/NHL_2017_skaters.html") %>%
  html_table(fill = TRUE)

nhl_2018_raw <- read_html("https://www.hockey-reference.com/leagues/NHL_2018_skaters.html") %>%
  html_table(fill = TRUE)

nhl_2019_raw <- read_html("https://www.hockey-reference.com/leagues/NHL_2019_skaters.html") %>%
  html_table(fill = TRUE)

nhl_2020_raw <- read_html("https://www.hockey-reference.com/leagues/NHL_2020_skaters.html") %>%
  html_table(fill = TRUE)

nhl_2021_raw <- read_html("https://www.hockey-reference.com/leagues/NHL_2021_skaters.html") %>%
  html_table(fill = TRUE)

nhl_2022_raw <- read_html("https://www.hockey-reference.com/leagues/NHL_2022_skaters.html") %>%
  html_table(fill = TRUE)

nhl_2023_raw <- read_html("https://www.hockey-reference.com/leagues/NHL_2023_skaters.html") %>%
  html_table(fill = TRUE)


# Clean each year

nhl_2013_clean <- nhl_2013_raw[[1]] %>%    #pull first table
  clean_names() %>%        #clean names
  filter(x_2 != "Player") %>%       #remove repeated header rows
  transmute(player = x_2, season = 2013, team = x_4, position = x_5,
            games = as.numeric(x_6), raw_performance = as.numeric(scoring_3),
            league = "NHL", performance_metric = "Points")

nhl_2014_clean <- nhl_2014_raw[[1]] %>%
  clean_names() %>%
  filter(x_2 != "Player") %>%
  transmute(player = x_2, season = 2014, team = x_4, position = x_5,
            games = as.numeric(x_6), raw_performance = as.numeric(scoring_3),
            league = "NHL", performance_metric = "Points")

nhl_2015_clean <- nhl_2015_raw[[1]] %>%
  clean_names() %>%
  filter(x_2 != "Player") %>%
  transmute(player = x_2, season = 2015, team = x_4, position = x_5,
            games = as.numeric(x_6), raw_performance = as.numeric(scoring_3),
            league = "NHL", performance_metric = "Points")

nhl_2016_clean <- nhl_2016_raw[[1]] %>%
  clean_names() %>%
  filter(x_2 != "Player") %>%
  transmute(player = x_2, season = 2016, team = x_4, position = x_5,
            games = as.numeric(x_6), raw_performance = as.numeric(scoring_3),
            league = "NHL", performance_metric = "Points")

nhl_2017_clean <- nhl_2017_raw[[1]] %>%
  clean_names() %>%
  filter(x_2 != "Player") %>%
  transmute(player = x_2, season = 2017, team = x_4, position = x_5,
            games = as.numeric(x_6), raw_performance = as.numeric(scoring_3),
            league = "NHL", performance_metric = "Points")

nhl_2018_clean <- nhl_2018_raw[[1]] %>%
  clean_names() %>%
  filter(x_2 != "Player") %>%
  transmute(player = x_2, season = 2018, team = x_4, position = x_5,
            games = as.numeric(x_6), raw_performance = as.numeric(scoring_3),
            league = "NHL", performance_metric = "Points")

nhl_2019_clean <- nhl_2019_raw[[1]] %>%
  clean_names() %>%
  filter(x_2 != "Player") %>%
  transmute(player = x_2, season = 2019, team = x_4, position = x_5,
            games = as.numeric(x_6), raw_performance = as.numeric(scoring_3),
            league = "NHL", performance_metric = "Points")

nhl_2020_clean <- nhl_2020_raw[[1]] %>%
  clean_names() %>%
  filter(x_2 != "Player") %>%
  transmute(player = x_2, season = 2020, team = x_4, position = x_5,
            games = as.numeric(x_6), raw_performance = as.numeric(scoring_3),
            league = "NHL", performance_metric = "Points")

nhl_2021_clean <- nhl_2021_raw[[1]] %>%
  clean_names() %>%
  filter(x_2 != "Player") %>%
  transmute(player = x_2, season = 2021, team = x_4, position = x_5,
            games = as.numeric(x_6), raw_performance = as.numeric(scoring_3),
            league = "NHL", performance_metric = "Points")

nhl_2022_clean <- nhl_2022_raw[[1]] %>%
  clean_names() %>%
  filter(x_2 != "Player") %>%
  transmute(player = x_2, season = 2022, team = x_4, position = x_5,
            games = as.numeric(x_6), raw_performance = as.numeric(scoring_3),
            league = "NHL", performance_metric = "Points")

nhl_2023_clean <- nhl_2023_raw[[1]] %>%
  clean_names() %>%
  filter(x_2 != "Player") %>%
  transmute(player = x_2, season = 2023, team = x_4, position = x_5,
            games = as.numeric(x_6), raw_performance = as.numeric(scoring_3),
            league = "NHL", performance_metric = "Points")

# Combine all years
nhl_clean <- bind_rows(
  nhl_2013_clean,
  nhl_2014_clean,
  nhl_2015_clean,
  nhl_2016_clean,
  nhl_2017_clean,
  nhl_2018_clean,
  nhl_2019_clean,
  nhl_2020_clean,
  nhl_2021_clean,
  nhl_2022_clean,
  nhl_2023_clean
)

#Adjust for players who played for multiple teams to only take their total season stast
nhl_clean <- nhl_clean %>%
  group_by(player, season) %>%
  filter(team == "TOT" | n() == 1) %>%
  ungroup()
```

#NHL salary data

\`\`\`
